---
title: "L4_pulldown_proteomics_analysis_V001"
output: html_document
date: "2026-03-14"
---



```{r}
# Load packages
library(tidyverse)
library(limma)
library(DEqMS)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(scales)# for label_scientific() in dotblots
library(msImpute)#For imputing 
```

#Import the data 
Get data from the pre-processed file, where headers for MS1 and MS2 columns have been modified to include 
only the sample and replicate info and columns indicating nr of detected peptides have been removed.
```{r}
data <- read.csv2("/Users/mikelre/Documents/project_05_L4_pulldown_&_Kinetics/L4_Pulldown/L4_pulldown_proteomics/proteomics_raw_data/proteomics_raw_data_preprocessed/CPMSF_GPPF-MRE-280_CF280_proteotypic_protein-pivot_V002.csv", check.names = FALSE, sep = ";")
```

#Import CrapOme dataframe 
```{r}
CrapOme_df <- read.csv2("/Users/mikelre/Documents/project_05_L4_pulldown_&_Kinetics/L4_Pulldown/L4_pulldown_proteomics/CrapOme/CrapOme_protein_list_V001.csv", check.names = FALSE, sep = ";")

CrapOme_df$GENE <- sub("_HUMAN$", "", CrapOme_df$GENE)

```

#PCA QC
```{r}
#Generate expression matrix
quant_cols_PCA <- c(
"1a_MS2","1b_MS2","1c_MS2",
"2a_MS2","2b_MS2","2c_MS2",
"3a_MS2","3b_MS2","3c_MS2",
"4a_MS2","4b_MS2","4c_MS2"
)

expr_PCA <- data[, quant_cols_PCA]
rownames(expr_PCA) <- data$PG.ProteinGroups

#First transform dataframe valiues to numeric:
# identify MS2 columns
ms2_cols <- grep("_MS2$", names(expr_PCA), value = TRUE)

# convert to numeric
expr_PCA[ms2_cols] <- lapply(expr_PCA[ms2_cols],
                         function(x) as.numeric(as.character(x)))


#Log2 transform
expr_log2_PCA <- log2(expr_PCA)

#Remove proteins with many missing values
keep <- rowSums(!is.na(expr_log2_PCA)) >= 3 # keep only proteins quantified (non-NA) in ≥x samples
expr_log2_PCA <- expr_log2_PCA[keep,]

#Inpute missing valuies for PCA only. PCA requires complete data, so temporarily fill missing values.Simple approach: replace NA with column medians.
#would this be better with row medians?

expr_pca <- expr_log2_PCA

for(i in 1:ncol(expr_pca)){
  expr_pca[is.na(expr_pca[,i]), i] <- median(expr_pca[,i], na.rm = TRUE)
}

#Run PCA
pca_data <- prcomp(t(expr_pca), scale. = TRUE)

# Get sample PC scores (PC coordinates for samples)
pca_df <- as.data.frame(pca_data$x)  # samples x PCs

#Add metadata to pca_df
sample_group <- data.frame(sample = quant_cols_PCA, 
                            group = c("TT+L4", "TT+L4", "TT+L4", 
                                      "L4", "L4", "L4", 
                                      "TT+Flag", "TT+Flag", "TT+Flag",
                                      "Flag", "Flag", "Flag") 
                            )

pca_df$group <- sample_group$group


# Calculate variance explained for labels
percentVar <- round(100 * (pca_data$sdev^2 / sum(pca_data$sdev^2)))


# Create PCA plot
pca_plot <- ggplot(pca_df, aes(x=PC1, y=PC2, color = group, label = group)) +
  geom_point(size=3) +
  geom_text(hjust = 0.5, vjust = -0.9, size = 4) +  # Adds labels above points
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal() +
  ggtitle("PCA of log2 transformed MS2 intensities") +
  theme(plot.title = element_text(hjust = 0.5))

print(pca_plot)
```

#Extract genes loadings
```{r}
# Extract gene loadings (contributions to PCs)
gene_loadings <- pca_data$rotation  # genes x PCs

#Get top contributing genes to PC1 and PC2. We do this by selecting the top X genes with highest absolute value of loading
ranked_genes_PC1 <- sort(abs(gene_loadings[, 1]), decreasing = TRUE)
ranked_genes_PC2 <- sort(abs(gene_loadings[, 2]), decreasing = TRUE)

# Display top 20 gene names contributing to PC1 and PC2
top_X_genes_PC1 <- names(head(ranked_genes_PC1, 100))
top_X_genes_PC2 <- names(head(ranked_genes_PC2, 50))

#Write csv files (useful for string analysis )
write.csv(top_X_genes_PC1, file = "Top_genes_PC1.csv", row.names = FALSE, quote = FALSE)
write.csv(top_X_genes_PC2, file = "Top_genes_PC2.csv", row.names = FALSE, quote = FALSE)

```

#Extract Quantification matrix

"1a_MS2","1b_MS2","1c_MS2", -> These samples are TT+L4
"2a_MS2","2b_MS2","2c_MS2", -> These samples are L4
"3a_MS2","3b_MS2","3c_MS2", -> These samples are TT+Flag
"4a_MS2","4b_MS2","4c_MS2" -> These samples are Flag
```{r}
quant_cols <- c(
"1a_MS2","1b_MS2","1c_MS2",
"3a_MS2","3b_MS2","3c_MS2"
)

expr <- data[,quant_cols]

#Set row names to proteins. We can change this to give gene name!
rownames(expr) <- data$PG.ProteinGroups


```



#Log2 transformation of the data
```{r}

#Fairst transform dataframe valiues to numeric:
# identify MS2 columns
ms2_cols <- grep("_MS2$", names(expr), value = TRUE)

# convert to numeric
expr[ms2_cols] <- lapply(expr[ms2_cols],
                         function(x) as.numeric(as.character(x)))

#Perform Log2 transformation.
expr_log2 <- log2(expr)

#Check if the median is centered
boxplot(expr_log2,las=2,main= "Median of samples. Not normalized. Visible test-control effect")

```

## Median centering: 
Proteomics data from mass spectrometry often shows systematic shifts due to run-to-run variations in instrument performance or sample preparation. Median centering scales each batch or sample so their medians match a common reference (e.g., overall dataset median), minimizing these artifacts without assuming normality

Median normalization aligns the centers of the data distributions. If your pull-down is successful, the test sample will have high-intensity, specific binders that the control lacks. Centering forces the median of the test sample down and the control up, potentially masking true biological differences

The goal of a pull-down experiment is to identify specific binders (present in the test sample) versus non-specific background binders (present in both). Median centering assumes that the majority of proteins remain constant across samples, which is not true in a pull-down where the test sample is intentionally highly enriched for specific proteins. So, It is not appropiate to do median centering in thios case!

Common aproach we can follow: Subtract the intensity of proteins in the control from the test sample to determine specific binders




#Cleaning step 
Fom test samples, remove all proteins that have NaN value in 2 or 3 replicates (in test sample)
```{r}
test_cols <- c("1a_MS2","1b_MS2","1c_MS2")

keep <- rowSums(!is.na(expr_log2[,test_cols])) >= 3 # keep only proteins quantified (non-NA) in ≥x samples. If I select 3, I do not allow any missing value in the test samples. Increase the confidence of detection and reduce the noise!

expr_log2 <- expr_log2[keep,]
data <- data[keep,]
```

#QC of Log2 transformation with 
```{r}
#Calculate the average intensity of the true interactors
# Columns to use
cols <- c("1a_MS2", "1b_MS2", "1c_MS2")

# Create summary dataframe with averaged intensityes
expr_log2_average_intensity <- data.frame(
  Mean_MS2 = rowMeans(true_interactors[, cols], na.rm = TRUE),
  SD_MS2 = apply(true_interactors[, cols], 1, sd, na.rm = TRUE)
)
```


#Taking the true positives 
```{r}
# Define columns (you already have this)
bait_cols <- grep("^1[abc]_MS2$", colnames(expr_log2), value = TRUE)
control_cols <- grep("^3[abc]_MS2$", colnames(expr_log2), value = TRUE)

# Condition 1: present in ALL bait samples
present_in_bait <- rowSums(!is.na(expr_log2[, bait_cols])) == length(bait_cols)#Keep proteins detected in all bait samples

# Condition 2: missing in ALL control samples
missing_in_control <- rowSums(!is.na(expr_log2[, control_cols])) == 0 #Keep proteins detected in all bait samples

# Combine conditions
true_interactors <- expr_log2[present_in_bait & missing_in_control, ]
#Transform protein names to gene names using the data df
rownames(true_interactors) <- data$PG.Genes[match(rownames(true_interactors), data$PG.ProteinGroups)]


#Calculate the average intensity of the true interactors
# Columns to use
cols <- c("1a_MS2", "1b_MS2", "1c_MS2")

# Create summary dataframe with averaged intensityes
true_interactors_average_intensity <- data.frame(
  Mean_MS2 = rowMeans(true_interactors[, cols], na.rm = TRUE),
  SD_MS2 = apply(true_interactors[, cols], 1, sd, na.rm = TRUE)
)

#sort the dataframe to get highest MS2 intensities oin top
true_interactors_average_intensity <- true_interactors_average_intensity %>%
  arrange(desc(Mean_MS2))

#Clean the list with CrapOme database
true_interactors_average_intensity_filtered <- true_interactors_average_intensity[
  !rownames(true_interactors_average_intensity) %in% CrapOme_df$GENE, 
]

#sort the dataframe to get highest MS2 intensities oin top
true_interactors_average_intensity_filtered <- true_interactors_average_intensity_filtered %>%
  arrange(desc(Mean_MS2))

hist(true_interactors_average_intensity_filtered$Mean_MS2,
     main = "Distribution of Mean MS2 Intensities",
     xlab = "Mean MS2 Intensity",
     ylab = "Frequency")

write.csv(as.data.frame(true_interactors_average_intensity_filtered), "true_interactors_average_intensity_filtered_TTL4vsTTFLAG.csv", row.names = TRUE)

#check which genes were filtered out
contaminants_in_true_interactors <- true_interactors_average_intensity[
  rownames(true_interactors_average_intensity) %in% CrapOme_df$GENE, 
]

```



#Define experimental design
```{r}
group <- factor(c(
"Test","Test","Test",
"Control","Control","Control"
))

design <- model.matrix(~0 + group)
colnames(design) <- levels(group)
```


#Fit Limma model 
```{r}
fit <- lmFit(expr_log2, design)

#Introduce comparissons
contrast <- makeContrasts(Test-Control, levels=design)

fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2)

```


#Prepare DEqMS Peptide Counts
DEqMS needs number of quantified peptides per protein. For this we use:
PG.NrOfPrecursorsIdentified_Experiment_wide_ -> Column in excel file (headers modified)

```{r}
pepcount <- data$PG.NrOfPrecursorsIdentified_Experiment_wide_
names(pepcount) <- rownames(expr_log2)
fit2$count <- pepcount

#QC: If this returns FALSE, the counts are not aligned with the proteins.
all(rownames(expr_log2) == names(pepcount))
#QC: All the values resulting below shold be equal
nrow(expr_log2)
length(pepcount)
length(fit2$count)
```


#Run DEqMS
```{r}
#Run DEqMS
fit3 <- spectraCounteBayes(fit2)

#Extract DEqMS resuls
results <- outputResult(fit3)
```


#Merge Protein Metadata
```{r}
results$Protein <- rownames(results)

results <- left_join(results,
                     data[,c("PG.ProteinGroups","PG.Genes","PG.ProteinDescriptions")],
                     by=c("Protein"="PG.ProteinGroups"))
```



#QC that the Limma model worked
```{r}
VarianceScatterplot(fit3)
VarianceBoxplot(fit3)
```

The red line is slightly higher at intermediate counts and then gently slopes down at higher counts; importantly, it is smooth and not wildly oscillating, and variance clearly does not explode at high counts. That’s perfectly acceptable: DEqMS has captured the modest dependence of variance on evidence count, and will now assign smaller P‑values to proteins with more evidence at the same fold‑change.

I would be concerned only if:

The red line were extremely wiggly (overfitting),

Or strongly increasing with counts (suggesting a modeling issue),

Or based on very few distinct count values (e.g. you accidentally passed a constant count vector).


#Downstream Analysis 

##Volcano plot
```{r}
adj.P.Val_value <- 0.001
logFC_value <- 2.5

results <- results %>%
  mutate(significance = case_when(
    adj.P.Val < adj.P.Val_value & logFC > logFC_value  ~ "Upregulated",
    adj.P.Val < adj.P.Val_value & logFC < -logFC_value ~ "Downregulated",
    TRUE ~ "Not significant"
  ))

ggplot(results, aes(logFC, -log10(adj.P.Val), color = significance)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_vline(xintercept = c(-logFC_value, logFC_value), linetype = "dashed") +
  geom_hline(yintercept = -log10(adj.P.Val_value), linetype = "dashed") +
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not significant" = "grey70")) +  
  geom_text_repel(data = subset(results, abs(logFC) > 5 & -log10(adj.P.Val) > 3),
                  aes(label = PG.Genes),  # Remove redundant x/y/color from aes()
                  color = "black",
                  max.overlaps = 15, size = 3) +
  theme_minimal() +
  ggtitle("TT+L4 vs TT") +
  labs(color = "Category")

```

#Extract upregulated and downregulated genes
```{r}
up_genes <- results$PG.Genes[results$sig == "Upregulated"] 
lapply(up_genes, function(x) x[!is.na(x)])
length(up_genes)

down_genes <- results$PG.Genes[results$sig == "Downregulated"]
length(down_genes)
```


#Filter results to remove genes in CrapOme database
```{r}
# Remove rows from df1 where 'gene' column matches ANY value in df2$gene
results_filtered <- results[!results$PG.Genes %in% CrapOme_df$GENE, ]

nrow(results)
nrow(results_filtered)
```

#Volcano plot after filtering
```{r}
adj.P.Val_value <- 0.001
logFC_value <- 2.5

results_filtered <- results_filtered %>%
  mutate(significance = case_when(
    adj.P.Val < adj.P.Val_value & logFC > logFC_value  ~ "Upregulated",
    adj.P.Val < adj.P.Val_value & logFC < -logFC_value ~ "Downregulated",
    TRUE ~ "Not significant"
  ))

ggplot(results_filtered, aes(logFC, -log10(adj.P.Val), color = significance)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_vline(xintercept = c(-logFC_value, logFC_value), linetype = "dashed") +
  geom_hline(yintercept = -log10(adj.P.Val_value), linetype = "dashed") +
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not significant" = "grey70")) +  
  geom_text_repel(data = subset(results_filtered, abs(logFC) > 5 & -log10(adj.P.Val) > 3),
                  aes(label = PG.Genes),  # Remove redundant x/y/color from aes()
                  color = "black",
                  max.overlaps = 15, size = 3) +
  #coord_cartesian(ylim = c(0, 10)) +   # Changing Y axis limmits
  theme_minimal() +
  ggtitle("TT+L4 vs TT. Filtered with CrapOme") +
  labs(color = "Category")

```


#Extract results of up and downregulated proteins
```{r}

up_genes_post_filtration <- results_filtered$PG.Genes[results_filtered$sig == "Upregulated"] 
lapply(up_genes_post_filtration, function(x) x[!is.na(x)]) #For each element in the list, remove all NA values
length(up_genes_post_filtration)

#Write a csv file 
write.csv(as.data.frame(up_genes_post_filtration), "Up_genes_post_filtration_TTL4vsTT.csv", row.names = TRUE)

down_genes_post_filtration <- results_filtered$PG.Genes[results_filtered$sig == "Downregulated"]
lapply(down_genes_post_filtration, function(x) x[!is.na(x)]) #For each element in the list, remove all NA values
length(down_genes_post_filtration)

#Write a csv file 
write.csv(as.data.frame(down_genes_post_filtration), "Down_genes_post_filtration_TTL4vsTT.csv", row.names = TRUE)

```






#Venn diagrams.
```{r}

# Sample data (replace with your actual vectors)
v1 <- top_X_genes_PC1
v2 <- top_X_genes_PC2
#v3 <- pseudogenes
#v3 <- KEGG_pathway_contributing_genes2

vec_list <- list(v1, v2)

venn_plot <- ggVennDiagram(
  vec_list,
  category.names = c("PC1", "PC2")
) +
  scale_fill_gradient(low = "#E0F7FA", high = "#FFCDD2") +
  theme_void() +
  labs(title = "Comparison of Gene Sets") +
  theme(
    text = element_text(family = "sans", size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  # centered title
    plot.margin = margin(10, 10, 10, 10),   # more margin so labels aren’t cut
    legend.position = "none"   
  ) +
  coord_equal() +  # keep circles round
  scale_x_continuous(expand = expansion(mult = 0.3)) +  # add space left/right
  scale_y_continuous(expand = expansion(mult = 0.2))    # add space top/bottom


# Display
venn_plot
```


















#Gene ontology analysis
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

#The GO is organized in three aspects: Molecular Function (MF), Cellular Component (CC), and Biological Process (BP)
      ontology_term <- "BP" #Change gere which ontology term I want to use for the analysis

enrichGO(gene = up_genes_post_filtration, OrgDb = org.Hs.eg.db, ont = ontology_term, keyType = "SYMBOL")


# genes_in_clusterX contains gene symbols
gene_symbols <- as.character(up_genes_post_filtration)

# Convert to Entrez IDs using org.Hs.eg.db
gene_entrez <- bitr(gene_symbols, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

print(dim(gene_entrez))
# Print the first few mapped genes
head(gene_entrez)

# GO enrichment without log2FC
go_results <- enrichGO(gene = gene_entrez$ENTREZID,  #Entrez IDs
                       OrgDb = org.Hs.eg.db,  # Specify the organism database
                       keyType = "ENTREZID",  # Specify key type
                       ont = ontology_term,  # Choose ontology: "BP" (Biological Process), "CC" (Cellular Component), or "MF" (Molecular Function)
                       pvalueCutoff = 0.05,  # Adjust p-value cutoff
                       qvalueCutoff = 0.05)  # Adjust q-value cutoff

# View results
(as.data.frame(go_results))

# Convert GO enrichment results to data frame
go_df <- as.data.frame(go_results)

# Save to CSV
#write.csv(go_df, file = "/Users/mikelre/Documents/project_01_processing_rna/nf_core/results_salmon_human_and_Virus/go_enrichment_results.csv", row.names = FALSE)


##Create a dot blot with GO analysis results

# X-axis: Gene ratio (by default)
# GeneRatio = (genes from my list in the term) ÷ (total genes submitted).
# It answers: “What proportion of my interesting genes are annotated with this GO term?”
# Y-axis: GO terms
# Dot size: Number of genes
# Dot color: Adjusted p-value (FDR)

#Number of ontologies to plot;Generate dot plot (top X categories)
number_ontologies_toplot <- 5

#recent enrichplot versions (≥1.18), the argument orderBy = "Count" is actually ignored if x = "GeneRatio" is set (default). Therefore we have to subset and reorder the results dataframe manually:

# take top N pathways by Count
top_df <- go_df %>%
  arrange(desc(Count)) %>%
  slice_head(n = number_ontologies_toplot)

# make Description ordered by Count (so y-axis respects it)
top_df$Description <- fct_reorder(top_df$Description, top_df$Count)

# rebuild an enrichResult-like object for plotting
go_results@result <- top_df

p <- dotplot(go_results, showCategory = number_ontologies_toplot, x = "GeneRatio") +
  scale_fill_continuous(labels = label_scientific(), low = "navy", high = "lightblue") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  ) +
  labs(
    x = "Gene Ratio",
    y = "",
    title = 
    #title = sprintf("Cluster %s/%s - Top %s enriched GO-%s Pathways", cluster_nr, number_of_clusters, number_ontologies_toplot, ontology_term)
  )


# Print plot
print(p)

```

