---
title: "L4_pulldown_proteomics_analysis_V001"
output: html_document
date: "2026-03-14"

V003: In this verssion I included imputation with msImpute from the begining of the pipeline.
---



```{r}
# Load packages
library(tidyverse)
library(limma)
library(DEqMS)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(scales)# for label_scientific() in dotblots
library(msImpute)#For imputing 
```

#Import the data 
Get data from the pre-processed file, where headers for MS1 and MS2 columns have been modified to include 
only the sample and replicate info and columns indicating nr of detected peptides have been removed.
```{r}
data <- read.csv2("/Users/mikelre/Documents/project_05_L4_pulldown_&_Kinetics/L4_Pulldown/L4_pulldown_proteomics/proteomics_raw_data/proteomics_raw_data_preprocessed/CPMSF_GPPF-MRE-280_CF280_proteotypic_protein-pivot_V002.csv", check.names = FALSE, sep = ";")
```

#Import CrapOme dataframe 
```{r}
CrapOme_df <- read.csv2("/Users/mikelre/Documents/project_05_L4_pulldown_&_Kinetics/L4_Pulldown/L4_pulldown_proteomics/CrapOme/CrapOme_protein_list_V001.csv", check.names = FALSE, sep = ";")

CrapOme_df$GENE <- sub("_HUMAN$", "", CrapOme_df$GENE)

```

#Log2 transform MS2 intensities 
```{r}
#Generate expression matrix
quant_cols <- c(
"1a_MS2","1b_MS2","1c_MS2",
"2a_MS2","2b_MS2","2c_MS2",
"3a_MS2","3b_MS2","3c_MS2",
"4a_MS2","4b_MS2","4c_MS2"
)

expr <- data[, quant_cols]
rownames(expr) <- data$PG.ProteinGroups

#First transform dataframe valiues to numeric:
# identify MS2 columns
ms2_cols <- grep("_MS2$", names(expr), value = TRUE)

# convert to numeric
expr[ms2_cols] <- lapply(expr[ms2_cols],
                         function(x) as.numeric(as.character(x)))


#Log2 transform
expr_log2 <- log2(expr)
#QC
boxplot(expr_log2,las=2,main= "Median of samples. Log2 transformed")

```

#------------------------------------------
#Inputing to remove NA. Imputing a very low intensity value is a commonly accepted, though not always optimal, method for handling NA values in pulldown (e.g., co-IP) proteomics experiments. In mass spectrometry (MS)-based proteomics, missing values (NAs) are usually "Missing Not At Random" (MNAR)

Parameters for imputation:
mat_filt: Input matrix (rows = peptides/features, columns = samples) with NAs indicating missing intensities.

Parameter: method = "v2-mnar"

  "v2-mnar" specifies the imputation method.
  
  In msImpute, this refers to missing not at random (MNAR) imputation
  
  MNAR assumes that missing values depend on the true abundance (e.g., low-abundance proteins are more likely to be missing).    We know this is true because we are pre-selecting proteins that have presence in all bait replicates, so we expect to have     missing values in the control samples
  
  The method estimates the missing values based on this assumption, often drawing from a down-shifted        Gaussian distribution.

Parameter group

  group specifies sample groupings (bait vs. control)
  
  Imputation can be done within groups so that the distribution of imputed values respects the differences between biological    conditions.
##Prepare data for imputation
```{r}
# =============================
# 0. Library
# =============================
library(msImpute)

# =============================
# 1. Prepare data
# =============================
mat <- as.matrix(expr_log2)

# Replace NaN with NA
mat[is.nan(mat)] <- NA

# =============================
# 2. Define groups
# =============================
bait_cols_AAV_context <- grep("^1[abc]_MS2$", colnames(mat), value = TRUE)
bait_cols_cell_context <- grep("^3[abc]_MS2$", colnames(mat), value = TRUE)
control_cols <- grep("^2[abc]_MS2$|^4[abc]_MS2$", colnames(mat), value = TRUE)

# Combine all bait columns
bait_cols <- c(bait_cols_AAV_context, bait_cols_cell_context)

# Assign group labels
group <- ifelse(colnames(mat) %in% bait_cols, "bait", "control")

# =============================
# 3. Pulldown-aware filtering
# =============================
# Keep proteins observed in bait. In at least 3 replicates of AAV context or (|) cell context

keep <- (
  rowSums(!is.na(mat[, bait_cols_AAV_context])) >= 3 |
  rowSums(!is.na(mat[, bait_cols_cell_context])) >= 3
)

mat_filt <- mat[keep, ]

#Identify which proteins have less than 4 missing vcalues (required criteria for msImpute)
# Count non-missing values per row in mat_filt
non_missing_counts <- rowSums(!is.na(mat_filt))

# Identify rows with fewer than 4 non-missing values
low_obs <- non_missing_counts < 4

# Subset mat_filt, keeping **same columns** but only these low-observed rows
matfit_crap <- mat_filt[low_obs, , drop = FALSE]  # drop=FALSE keeps it as a dataframe/matrix


```

##Clean proteins that do not have more than 4 non missing values from matfilt
IT IS ONLY 1: Q92934
```{r}
# Keep only rows that are NOT in matfit_crap
mat_filt_clean <- mat_filt[!rownames(mat_filt) %in% rownames(matfit_crap), , drop = FALSE]

```


##Perform imputation
```{r}
# =============================
# 4. MNAR imputation
# =============================
expr_log2_imputed <- msImpute(
  mat_filt_clean,
  method = "v2-mnar",
  group = group
)

# =============================
# 5. Output
# =============================
# imputed matrix
expr_log2_imputed_df <- as.data.frame(expr_log2_imputed)

#Check if the median is centered
boxplot(expr_log2_imputed_df,las=2,main= "Median of samples. Imputed and Log2 transformed")
```

#------------------------------------------
#PCA with Imputed data
```{r}
#Run PCA
pca_data <- prcomp(t(expr_log2_imputed_df), scale. = TRUE)

# Get sample PC scores (PC coordinates for samples)
pca_df <- as.data.frame(pca_data$x)  # samples x PCs

#Add metadata to pca_df
sample_group <- data.frame(sample = quant_cols, 
                            group = c("TT+L4", "TT+L4", "TT+L4", 
                                      "L4", "L4", "L4", 
                                      "TT+Flag", "TT+Flag", "TT+Flag",
                                      "Flag", "Flag", "Flag") 
                            )

pca_df$group <- sample_group$group


# Calculate variance explained for labels
percentVar <- round(100 * (pca_data$sdev^2 / sum(pca_data$sdev^2)))


# Create PCA plot
pca_plot <- ggplot(pca_df, aes(x=PC1, y=PC2, color = group, label = group)) +
  geom_point(size=3) +
  geom_text(hjust = 0.5, vjust = -0.9, size = 4) +  # Adds labels above points
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal() +
  ggtitle("PCA of log2 transformed MS2 intensities (+ Imputation)") +
  theme(plot.title = element_text(hjust = 0.5))

print(pca_plot)
```

##Extract genes loadings
```{r}
# Extract gene loadings (contributions to PCs)
gene_loadings <- pca_data$rotation  # genes x PCs

#Get top contributing genes to PC1 and PC2. We do this by selecting the top X genes with highest absolute value of loading
ranked_genes_PC1 <- sort(abs(gene_loadings[, 1]), decreasing = TRUE)
ranked_genes_PC2 <- sort(abs(gene_loadings[, 2]), decreasing = TRUE)

# Display top 20 gene names contributing to PC1 and PC2
top_X_genes_PC1 <- names(head(ranked_genes_PC1, 100))
top_X_genes_PC2 <- names(head(ranked_genes_PC2, 50))

#Write csv files (useful for string analysis )
write.csv(top_X_genes_PC1, file = "Top_genes_PC1.csv", row.names = FALSE, quote = FALSE)
write.csv(top_X_genes_PC2, file = "Top_genes_PC2.csv", row.names = FALSE, quote = FALSE)

```

#------------------------------------------
#Extract Quantification matrix

"1a_MS2","1b_MS2","1c_MS2", -> These samples are TT+L4
"2a_MS2","2b_MS2","2c_MS2", -> These samples are L4
"3a_MS2","3b_MS2","3c_MS2", -> These samples are TT+Flag
"4a_MS2","4b_MS2","4c_MS2" -> These samples are Flag

HERE CHANGE QUANTIFICATION MATRIX TOI CHANGE COMPARISSONS!!!!
```{r}
quant_cols <- c(
"2a_MS2","2b_MS2","2c_MS2",
"4a_MS2","4b_MS2","4c_MS2"
)

expr_log2_imputed_df_for_DE <- expr_log2_imputed_df[,quant_cols]


#Check if the median is centered
boxplot(expr_log2_imputed_df_for_DE,las=2,main= "Median of samples. Not normalized. Visible test-control effect")

```









## Median centering: 
Proteomics data from mass spectrometry often shows systematic shifts due to run-to-run variations in instrument performance or sample preparation. Median centering scales each batch or sample so their medians match a common reference (e.g., overall dataset median), minimizing these artifacts without assuming normality

Median normalization aligns the centers of the data distributions. If your pull-down is successful, the test sample will have high-intensity, specific binders that the control lacks. Centering forces the median of the test sample down and the control up, potentially masking true biological differences

The goal of a pull-down experiment is to identify specific binders (present in the test sample) versus non-specific background binders (present in both). Median centering assumes that the majority of proteins remain constant across samples, which is not true in a pull-down where the test sample is intentionally highly enriched for specific proteins. So, It is not appropiate to do median centering in thios case!

Common aproach we can follow: Subtract the intensity of proteins in the control from the test sample to determine specific binders



#DEqMS 

##Define experimental design

```{r}
group <- factor(c(
"Test","Test","Test",
"Control","Control","Control"
))

design <- model.matrix(~0 + group)
colnames(design) <- levels(group)

```


##Fit Limma model 
```{r}
fit <- lmFit(expr_log2_imputed_df_for_DE, design)

#Introduce comparissons
contrast <- makeContrasts(Test-Control, levels=design)

fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2)

```


##Prepare DEqMS Peptide Counts
DEqMS needs number of quantified peptides per protein. For this we use:
PG.NrOfPrecursorsIdentified_Experiment_wide_ -> Column in excel file (headers modified)

```{r}
# Extract peptide counts in the order of expr_log2_imputed_df_for_DE
pepcount <- data$PG.NrOfPrecursorsIdentified_Experiment_wide_[
  match(
    rownames(expr_log2_imputed_df_for_DE),
    data$PG.ProteinGroups
  )
]

# Assign names to keep track
names(pepcount) <- rownames(expr_log2_imputed_df_for_DE)

fit2$count <- pepcount

#QC: If this returns FALSE, the counts are not aligned with the proteins.
all(rownames(expr_log2_imputed_df_for_DE) == names(pepcount))
#QC: All the values resulting below shold be equal
nrow(expr_log2_imputed_df_for_DE)
length(pepcount)
length(fit2$count)
```


##Run DEqMS
```{r}
#Run DEqMS
fit3 <- spectraCounteBayes(fit2)

#Extract DEqMS resuls
results <- outputResult(fit3)
```


##Merge Protein Metadata
```{r}
results$Protein <- rownames(results)

results <- left_join(results,
                     data[,c("PG.ProteinGroups","PG.Genes","PG.ProteinDescriptions")],
                     by=c("Protein"="PG.ProteinGroups"))
```



##QC that the Limma model worked
```{r}
VarianceScatterplot(fit3)
VarianceBoxplot(fit3)
```

The red line is slightly higher at intermediate counts and then gently slopes down at higher counts; importantly, it is smooth and not wildly oscillating, and variance clearly does not explode at high counts. That’s perfectly acceptable: DEqMS has captured the modest dependence of variance on evidence count, and will now assign smaller P‑values to proteins with more evidence at the same fold‑change.

I would be concerned only if:

The red line were extremely wiggly (overfitting),

Or strongly increasing with counts (suggesting a modeling issue),

Or based on very few distinct count values (e.g. you accidentally passed a constant count vector).

#------------------------------------------
#Downstream Analysis 
##Volcano plot
```{r}
adj.P.Val_value <- 0.001
logFC_value <- 2.5

results <- results %>%
  mutate(significance = case_when(
    adj.P.Val < adj.P.Val_value & logFC > logFC_value  ~ "Upregulated",
    adj.P.Val < adj.P.Val_value & logFC < -logFC_value ~ "Downregulated",
    TRUE ~ "Not significant"
  ))

ggplot(results, aes(logFC, -log10(adj.P.Val), color = significance)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_vline(xintercept = c(-logFC_value, logFC_value), linetype = "dashed") +
  geom_hline(yintercept = -log10(adj.P.Val_value), linetype = "dashed") +
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not significant" = "grey70")) +  
  geom_text_repel(data = subset(results, abs(logFC) > 5 & -log10(adj.P.Val) > 3),
                  aes(label = PG.Genes),  # Remove redundant x/y/color from aes()
                  color = "black",
                  max.overlaps = 15, size = 3) +
  theme_minimal() +
  ggtitle("L4-Flag vs Flag") +
  labs(color = "Category")

```

##Extract upregulated and downregulated genes
```{r}
up_genes <- results$PG.Genes[results$sig == "Upregulated"] 
lapply(up_genes, function(x) x[!is.na(x)])
length(up_genes)

down_genes <- results$PG.Genes[results$sig == "Downregulated"]
length(down_genes)
```


##Filter results to remove genes in CrapOme database
```{r}
# Remove rows from df1 where 'gene' column matches ANY value in df2$gene
results_filtered <- results[!results$PG.Genes %in% CrapOme_df$GENE, ]

nrow(results)
nrow(results_filtered)
```

###Volcano plot after filtering
```{r}
adj.P.Val_value <- 0.001
logFC_value <- 2.5

results_filtered <- results_filtered %>%
  mutate(significance = case_when(
    adj.P.Val < adj.P.Val_value & logFC > logFC_value  ~ "Upregulated",
    adj.P.Val < adj.P.Val_value & logFC < -logFC_value ~ "Downregulated",
    TRUE ~ "Not significant"
  ))

ggplot(results_filtered, aes(logFC, -log10(adj.P.Val), color = significance)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_vline(xintercept = c(-logFC_value, logFC_value), linetype = "dashed") +
  geom_hline(yintercept = -log10(adj.P.Val_value), linetype = "dashed") +
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not significant" = "grey70")) +  
  geom_text_repel(data = subset(results_filtered, abs(logFC) > 5 & -log10(adj.P.Val) > 3),
                  aes(label = PG.Genes),  # Remove redundant x/y/color from aes()
                  color = "black",
                  max.overlaps = 15, size = 3) +
  #coord_cartesian(ylim = c(0, 10)) +   # Changing Y axis limmits
  theme_minimal() +
  ggtitle("L4-Flag vs Flag. Filtered with CrapOme") +
  labs(color = "Category")

```


###Extract results of up and downregulated proteinsafter CrapOme filtration
```{r}

up_genes_post_filtration <- results_filtered$PG.Genes[results_filtered$sig == "Upregulated"] 
lapply(up_genes_post_filtration, function(x) x[!is.na(x)]) #For each element in the list, remove all NA values
length(up_genes_post_filtration)

#Write a csv file 
write.csv(as.data.frame(up_genes_post_filtration), "Up_genes_post_filtration_L4vsflag_IMPUTED.csv", row.names = TRUE)

down_genes_post_filtration <- results_filtered$PG.Genes[results_filtered$sig == "Downregulated"]
lapply(down_genes_post_filtration, function(x) x[!is.na(x)]) #For each element in the list, remove all NA values
length(down_genes_post_filtration)

#Write a csv file 
write.csv(as.data.frame(down_genes_post_filtration), "Down_genes_post_filtration_L4vsflaT_IMPUTED.csv", row.names = TRUE)

```


#------------------------------------------
#Ven diagrams Production vs Cells only

##Import interactors in both contexts

```{r}

interactors_production <- read.csv2("/Users/mikelre/Documents/project_05_L4_pulldown_&_Kinetics/L4_Pulldown/L4_pulldown_proteomics/code/Up_genes_post_filtration_TTL4vsTT_IMPUTED.csv", check.names = FALSE, sep = ";")

interactors_cells_only <- read.csv2("/Users/mikelre/Documents/project_05_L4_pulldown_&_Kinetics/L4_Pulldown/L4_pulldown_proteomics/code/Up_genes_post_filtration_L4vsflag_IMPUTED.csv", check.names = FALSE, sep = ";")

```


## Perform Venn diagrams.
```{r}

# Sample data (replace with your actual vectors)
v1 <- top_X_genes_PC1
v2 <- top_X_genes_PC2
#v3 <- pseudogenes
#v3 <- KEGG_pathway_contributing_genes2

vec_list <- list(v1, v2)

venn_plot <- ggVennDiagram(
  vec_list,
  category.names = c("PC1", "PC2")
) +
  scale_fill_gradient(low = "#E0F7FA", high = "#FFCDD2") +
  theme_void() +
  labs(title = "Comparison of Gene Sets") +
  theme(
    text = element_text(family = "sans", size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  # centered title
    plot.margin = margin(10, 10, 10, 10),   # more margin so labels aren’t cut
    legend.position = "none"   
  ) +
  coord_equal() +  # keep circles round
  scale_x_continuous(expand = expansion(mult = 0.3)) +  # add space left/right
  scale_y_continuous(expand = expansion(mult = 0.2))    # add space top/bottom


# Display
venn_plot
```




#------------------------------------------

#Gene ontology analysis
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

#The GO is organized in three aspects: Molecular Function (MF), Cellular Component (CC), and Biological Process (BP)
      ontology_term <- "BP" #Change gere which ontology term I want to use for the analysis

enrichGO(gene = up_genes_post_filtration, OrgDb = org.Hs.eg.db, ont = ontology_term, keyType = "SYMBOL")


# genes_in_clusterX contains gene symbols
gene_symbols <- as.character(up_genes_post_filtration)

# Convert to Entrez IDs using org.Hs.eg.db
gene_entrez <- bitr(gene_symbols, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

print(dim(gene_entrez))
# Print the first few mapped genes
head(gene_entrez)

# GO enrichment without log2FC
go_results <- enrichGO(gene = gene_entrez$ENTREZID,  #Entrez IDs
                       OrgDb = org.Hs.eg.db,  # Specify the organism database
                       keyType = "ENTREZID",  # Specify key type
                       ont = ontology_term,  # Choose ontology: "BP" (Biological Process), "CC" (Cellular Component), or "MF" (Molecular Function)
                       pvalueCutoff = 0.05,  # Adjust p-value cutoff
                       qvalueCutoff = 0.05)  # Adjust q-value cutoff

# View results
(as.data.frame(go_results))

# Convert GO enrichment results to data frame
go_df <- as.data.frame(go_results)

# Save to CSV
#write.csv(go_df, file = "/Users/mikelre/Documents/project_01_processing_rna/nf_core/results_salmon_human_and_Virus/go_enrichment_results.csv", row.names = FALSE)


##Create a dot blot with GO analysis results

# X-axis: Gene ratio (by default)
# GeneRatio = (genes from my list in the term) ÷ (total genes submitted).
# It answers: “What proportion of my interesting genes are annotated with this GO term?”
# Y-axis: GO terms
# Dot size: Number of genes
# Dot color: Adjusted p-value (FDR)

#Number of ontologies to plot;Generate dot plot (top X categories)
number_ontologies_toplot <- 10

#recent enrichplot versions (≥1.18), the argument orderBy = "Count" is actually ignored if x = "GeneRatio" is set (default). Therefore we have to subset and reorder the results dataframe manually:

# take top N pathways by Count
top_df <- go_df %>%
  arrange(desc(Count)) %>%
  slice_head(n = number_ontologies_toplot)

# make Description ordered by Count (so y-axis respects it)
top_df$Description <- fct_reorder(top_df$Description, top_df$Count)

# rebuild an enrichResult-like object for plotting
go_results@result <- top_df

p <- dotplot(go_results, showCategory = number_ontologies_toplot, x = "GeneRatio") +
  scale_fill_continuous(labels = label_scientific(), low = "navy", high = "lightblue") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  ) +
  labs(
    x = "Gene Ratio",
    y = "",
    title = sprintf("Upregulated genes in L4vsFlag. Log2FC: %s  -log10(adj.p-val): %s - Top %s enriched GO-%s Ontologies", logFC_value, adj.P.Val_value, number_ontologies_toplot, ontology_term)
  )


# Print plot
print(p)

```


## Extract genes belonging to each GO term:
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

# Get result table
go_df <- as.data.frame(go_results)

# Choose the GO term you're interested in ID
term_ID <- "GO:1901987"  

# Find the row corresponding to that term
term_row <- go_df[go_df$ID == term_ID, ]

# Split the contributing Entrez IDs
entrez_ids <- unlist(strsplit(term_row$geneID, split = "/"))

# Convert Entrez IDs to gene symbols (returns a named vector)
gene_symbols <- mapIds(
  org.Hs.eg.db,
  keys = entrez_ids,
  column = "SYMBOL",
  keytype = "ENTREZID",
  multiVals = "first"
)

# Convert to a clean character vector (unnamed, no NAs)
GO_contributing_genes_vector <- na.omit(unname(gene_symbols))

# View the result
print(GO_contributing_genes_vector)
write.csv(data.frame(Gene = GO_contributing_genes_vector), file = "GO_contributing_genes.csv", row.names = FALSE)


```

